---
title: "Processing and Clustering of Simulated Data Using the Rsearch Package"
author: "Cassandra Stamsaas"
date: "2026-07-23"
output: 
  html_document:
    code_folding: show
    number_sections: false
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 4
    df_print: paged
editor_options: 
  chunk_output_type: console
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Required Packages

```{r, message=FALSE, warning=FALSE}
library(forcats)
library(mclust)
library(microseq)
library(Rsearch)
library(tidyverse)
```

## Global Settings

```{r}
setwd("/path/to/working/directory") # Path to your working directory
```

## Prepare Directories and Paths

```{r}
path                <- "1000_species_10_dynamic"          # Define dataset to process
fastq.folder        <- file.path(path, "fastq")           # Directory containing the raw FASTQ files
fastq_filt.folder   <- file.path(path, "fastq_filtered")  # Directory containing quality-filtered FASTQ files
fasta_merged.folder <- file.path(path, "fasta_merged")    # Directory containing merged FASTA files
rsearch_dir         <- file.path(path, "rsearch_results") # Directory containing Rsearch output
fasta.folder        <- file.path(rsearch_dir, "fasta")    # Directory containing dereplicated FASTA files
metadata.file       <- file.path(path, "metadata.tsv")    # Path to the metadata file

# ------------------
# cluster_size paths
# ------------------

cluster_size_folder                  <- file.path(rsearch_dir, 
                                                  "cluster_size")     # Directory containing cluster_size results
dominant_species_folder_cluster_size <- file.path(cluster_size_folder, 
                                                  "dominant_species") # Directory containing dominant-species assignments for each cluster
abundance_folder_cluster_size        <- file.path(cluster_size_folder, 
                                                  "abundance_tables") # Directory containing observed and ground truth abundance tables
noise_seqs_cluster_size              <- file.path(cluster_size_folder, 
                                                  "noise_seqs")       # Directory containing sequences classified as noise

# --------------------
# cluster_unoise paths
# --------------------

cluster_unoise_folder                  <- file.path(rsearch_dir, 
                                                    "cluster_unoise_8") # Directory containing cluster_unoise results

dominant_species_folder_cluster_unoise <- file.path(cluster_unoise_folder, 
                                                    "dominant_species") # Directory containing dominant-species assignments for each cluster

abundance_folder_cluster_unoise        <- file.path(cluster_unoise_folder, 
                                                    "abundance_tables") # Directory containing observed and ground truth abundance tables

noise_seqs_cluster_unoise              <- file.path(cluster_unoise_folder, 
                                                    "noise_seqs")       # Directory containing sequences classified as noise

# ----------------------------
# Create required directories
# ----------------------------

if (!dir.exists(fastq_filt.folder)) {
  dir.create(fastq_filt.folder)
}

if (!dir.exists(fasta_merged.folder)) {
  dir.create(fasta_merged.folder)
}

if (!dir.exists(rsearch_dir)) {
  dir.create(rsearch_dir)
}

if (!dir.exists(fasta.folder)) {
  dir.create(fasta.folder)
}

# Create directories for cluster_size output
if (!dir.exists(cluster_size_folder)) {
  dir.create(cluster_size_folder)
}

if (!dir.exists(dominant_species_folder_cluster_size)) {
  dir.create(dominant_species_folder_cluster_size)
}

if (!dir.exists(abundance_folder_cluster_size)) {
  dir.create(abundance_folder_cluster_size)
}

if (!dir.exists(noise_seqs_cluster_size)) {
  dir.create(noise_seqs_cluster_size)
}

# Create directories for cluster_unoise output
if (!dir.exists(cluster_unoise_folder)) {
  dir.create(cluster_unoise_folder)
}

if (!dir.exists(dominant_species_folder_cluster_unoise)) {
  dir.create(dominant_species_folder_cluster_unoise)
}

if (!dir.exists(abundance_folder_cluster_unoise)) {
  dir.create(abundance_folder_cluster_unoise)
}

if (!dir.exists(noise_seqs_cluster_unoise)) {
  dir.create(noise_seqs_cluster_unoise)
}
```

## Metadata

```{r}
# Read metadata file and add columns for results
metadata.tbl <- read_tsv(metadata.file) %>% 
  select(-c(community_id, amplicon_file)) %>%
  mutate(dereplicated_fasta = NA) %>%
  rename(n_reads = readpairs) %>% 
  mutate(dataset = path) %>%
  mutate(n_reads_filtered = 0) %>% 
  mutate(n_reads_merged = 0) %>% 
  mutate(n_reads_dereplicated = 0) %>% 
  mutate(n_reads_wo_chimera = 0) %>% 
  mutate(n_clusters = 0) %>% 
  mutate(n_pure_clusters = 0) %>%
  mutate(n_perfect_clusters = 0) %>%
  mutate(average_purity_score = 0) %>%
  mutate(n_species = 0) %>% 
  mutate(n_species_filtered = 0) %>% 
  mutate(n_species_merged = 0) %>% 
  mutate(n_species_dereplicated = 0) %>% 
  mutate(n_species_wo_chimera = 0) %>% 
  mutate(n_species_pre_clustering = 0) %>% 
  mutate(n_species_represented = 0) %>% 
  mutate(n_species_with_multiple_clusters = 0) %>% 
  mutate(abundance_correlation = 0)
```

## Plot Base Quality

```{r}
for (i in 1:nrow(metadata.tbl)) {                                       # Iterate over samples in the metadata table
  
  R1.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R1_file[i])) # Read forward (R1) FASTQ reads for the current sample
  R2.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R2_file[i])) # Read reverse (R2) FASTQ reads for the current sample
  
  gg.obj <- plot_base_quality(
    R1.tbl, 
    R2.tbl, 
    show_overlap_box = TRUE,
    plot_title = paste("Sample:", metadata.tbl$sample_id[i])) # Create plot
  
  print(gg.obj)                                               # Visualize plot                 
  
  break # Remove this line to inspect base quality profiles for all samples
}
```

# Optimize Trimming 

## Optimize truncee_rate

```{r}
opt.tbl <- tibble(
  sample_id = metadata.tbl$sample_id,
  truncee_rate = 0,
  n_merged = 0) # Initialize a table for storing optimized truncation parameters and merging results

for (i in 1:nrow(metadata.tbl)) { # Iterate over samples in the metadata table
  
  R1.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R1_file[i])) # Read forward (R1) FASTQ reads for the current sample
  R2.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R2_file[i])) # Read reverse (R2) FASTQ reads for the current sample
  
  optimize_truncee_rate.tbl <- vs_optimize_truncee_rate(
    fastq_input = R1.tbl, 
    reverse = R2.tbl) # Run optimization function for truncee_rate
  
  print(attr(optimize_truncee_rate.tbl, "plot")) # Display the optimization results plot
  
  # -----------------------
  # Select optimal setting
  # -----------------------
  
  idx <- which.max(optimize_truncee_rate.tbl$merged_read_pairs) # Identify the truncee_rate yielding the highest number of merged read pairs
  
  opt.tbl$truncee_rate[i] <- 
    optimize_truncee_rate.tbl$truncee_rate_value[idx]           # Record the optimal truncee_rate for the current sample
  
  opt.tbl$n_merged[i] <- 
    optimize_truncee_rate.tbl$merged_read_pairs[idx]            # Record the corresponding number of successfully merged read pairs
}

save(
  opt.tbl,
  file = file.path(rsearch_dir, "opt.tbl.RData")) # Save optimized truncation parameters for downstream processing
```

# Trimming, Merging, Dereplication, and Chimera Detection

```{r}
load(file.path(rsearch_dir, "opt.tbl.RData")) # Load optimized truncee_rate values for each sample

for(i in 1:nrow(metadata.tbl)){ # Iterate over each sample in the metadata table
  
  message(paste0("Processing sample ", 
                 metadata.tbl$sample_id[i], 
                 " (", i, "/", nrow(metadata.tbl), ")"))
  
  R1.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R1_file[i])) # Read forward (R1) FASTQ reads for the current sample
  R2.tbl <- readFastq(file.path(fastq.folder, metadata.tbl$R2_file[i])) # Read reverse (R2) FASTQ reads for the current sample
  
  # -------------------------
  # Initial species richness
  # -------------------------
  
  df <- R1.tbl %>%
    mutate(species = Header %>% 
             str_extract("(?<=s:).+?(?=_\\d+_\\d+/)"))    # Extract the ground truth species label from each read header
  
  metadata.tbl$n_species[i] <- length(unique(df$species)) # Record the number of species present before read processing
  
  # --------------------
  # Trim and filter reads
  # --------------------
  
  cat("Trimming...", "\n")
  
  vs_fastx_trim_filt(
    R1.tbl, 
    R2.tbl,
    minlen = 20,                            # Minimum retained read length; default used by DADA2
    truncee_rate = opt.tbl$truncee_rate[i], # Apply the optimized truncation parameter for the current sample
    maxee_rate = 0.01,                      # Maximum expected-error rate used during truncee_rate optimization
    fastqout = file.path(
      fastq_filt.folder, 
      paste0(metadata.tbl$sample_id[i], "_R1_filtered.fastq")
    ),
    fastqout_rev = file.path(
      fastq_filt.folder, 
      paste0(metadata.tbl$sample_id[i], "_R2_filtered.fastq"))) # Trim and quality-filter paired-end reads and write filtered FASTQ files
  
  R1_filt.tbl <- readFastq(
    file.path(
      fastq_filt.folder, 
      paste0(metadata.tbl$sample_id[i], "_R1_filtered.fastq"))) # Read filtered forward reads
  
  R2_filt.tbl <- readFastq(
    file.path(
      fastq_filt.folder, 
      paste0(metadata.tbl$sample_id[i], "_R2_filtered.fastq"))) # Read filtered reverse reads
  
  metadata.tbl$n_reads_filtered[i] <- nrow(R1_filt.tbl) # Record the number of read pairs retained after filtering
  
  df <- R1_filt.tbl %>%
    mutate(species = Header %>% 
             str_extract("(?<=s:).+?(?=_\\d+_\\d+/)"))             # Extract ground truth species labels from filtered reads
  
  metadata.tbl$n_species_filtered[i] <- length(unique(df$species)) # Record the number of species represented after filtering
  
  # ----------------
  # Merge read pairs
  # ----------------
  
  cat("Merging...", "\n")
  
  vs_fastq_mergepairs(R1_filt.tbl, 
                      R2_filt.tbl,
                      fastaout = file.path(
                        fasta_merged.folder, 
                        paste0(metadata.tbl$sample_id[i], "_merged.fasta"))) # Merge overlapping forward and reverse reads and write merged sequences to FASTA
  
  merged.tbl <- readFasta(
    file.path(fasta_merged.folder, 
              paste0(metadata.tbl$sample_id[i], "_merged.fasta"))) # Read merged sequences for downstream processing
  
  metadata.tbl$n_reads_merged[i] <- nrow(merged.tbl) # Record the number of successfully merged read pairs
  
  df <- merged.tbl %>%
    mutate(species = Header %>% 
             str_extract("(?<=s:).+?(?=_\\d+_\\d+/)")) # Extract ground truth species labels from merged reads
  
  metadata.tbl$n_species_merged[i] <- length(unique(df$species)) # Record the number of species represented after merging
  
  # -------------
  # Dereplication
  # -------------
  
  cat("Dereplication...", "\n")
  
  derep.tbl <- vs_fastx_uniques(fastx_input = merged.tbl) # Dereplicate sequences
  
  n_derep_reads_total <- derep.tbl %>% 
    mutate(size = str_extract(Header, "(?<=;size=)\\d+")) %>% # Extract the read abundance encoded in each dereplicated FASTA header
    mutate(size = as.numeric(size)) %>%
    select(size) %>%
    sum()                                                     # Recover the total number of reads represented by dereplicated sequences
  
  metadata.tbl$n_reads_dereplicated[i] <- n_derep_reads_total # Record the total read count represented after dereplication
  
  df <- derep.tbl %>%
    mutate(species = Header %>% 
             str_extract("(?<=s:).+?(?=_\\d+_\\d+/)")) # Extract ground truth species labels from dereplicated sequences
  
  metadata.tbl$n_species_dereplicated[i] <- 
    length(unique(df$species)) # Record the number of species represented after dereplication
  
  # -----------------
  # Chimera detection
  # -----------------
  
  cat("Chimera detection...", "\n")
  
  derep_wo_chimera.tbl <- vs_uchime_denovo(derep.tbl) # Detect and remove chimeric sequences from the dereplicated dataset
  
  n_wo_chimera_reads_total <- derep_wo_chimera.tbl %>% 
    mutate(size = str_extract(Header, "(?<=;size=)\\d+")) %>% # Extract the abundance of each non-chimeric dereplicated sequence
    mutate(size = as.numeric(size)) %>%
    select(size) %>%
    sum()                                                     # Calculate the total number of reads retained after chimera removal
  
  metadata.tbl$n_reads_wo_chimera[i] <- n_wo_chimera_reads_total # Record the number of reads retained after chimera removal
  
  df <- derep_wo_chimera.tbl %>%
    mutate(species = Header %>% 
             str_extract("(?<=s:).+?(?=_\\d+_\\d+/)")) # Extract ground truth species labels from non-chimeric sequences
  
  metadata.tbl$n_species_wo_chimera[i] <- length(unique(df$species)) # Record the number of species represented after chimera removal
  
  # -----------------------------
  # Write processed FASTA output
  # -----------------------------
  
  writeFasta(
    derep_wo_chimera.tbl, 
    out.file = file.path(
      fasta.folder, 
      paste0(metadata.tbl$sample_id[i], "_dereplicated.fasta"))) # Write dereplicated, non-chimeric sequences to FASTA for downstream clustering
  
  metadata.tbl$dereplicated_fasta[i] <- 
    file.path(paste0(metadata.tbl$sample_id[i], "_dereplicated.fasta")) # Record the dereplicated FASTA filename in the metadata table
}

save(metadata.tbl, 
     file = file.path(rsearch_dir, "metadata.tbl.RData")) # Save the metadata table with read- and species-level processing statistics    
```

# cluster_size Clustering

```{r}
id      <- 0.99 # Sequence identity threshold used for clustering and noise-sequence mapping
minsize <- 2    # Minimum sequence abundance required for inclusion in initial clustering

load(file.path(rsearch_dir, "metadata.tbl.RData")) # Load metadata generated during read preprocessing

for (i in 1:nrow(metadata.tbl)) { # Iterate over each sample in the metadata table
  
  message(paste0("Processing sample ", 
                 metadata.tbl$sample_id[i], 
                 " (", i, "/", nrow(metadata.tbl), ")"))
  
  # ---------------------------
  # Load and parse input data
  # ---------------------------
  
  all.tbl <- readFasta(file.path(
    fasta.folder, metadata.tbl$dereplicated_fasta[i])) %>%                       # Read dereplicated sequences for current sample
    mutate(actual_cluster = Header,
           actual_cluster = str_match(actual_cluster, 
                                      "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>% # Extract the ground truth species label from each FASTA header
    mutate(size = str_extract(Header, "(?<=;size=)\\d+") %>% 
             as.numeric())                                                       # Parse sequence abundance (read count) from each dereplicated FASTA header
  
  metadata.tbl$n_species_pre_clustering[i] <- 
    length(unique(all.tbl$actual_cluster)) # Record the number of species represented immediately before clustering
  
  # ---------------------------------
  # Separate "bio" and noise sequences
  # ---------------------------------
  
  bio_seqs.tbl <- all.tbl %>%
    filter(size >= minsize) # Retain sequences meeting the minimum abundance threshold for initial clustering
  
  noise_seqs.tbl <- all.tbl %>%
    filter(size < minsize)  # Separate low-abundance sequences excluded from initial initial clustering
  
  
  # -----------------------
  # Clustering
  # -----------------------
  
  centroids_uc.tbl <- vs_cluster_size(
    bio_seqs.tbl, 
    uc = TRUE,
    sizeout = FALSE,
    id = id,
    size_column = TRUE,
    threads = 4) # Cluster "bio" sequences to find centroids and return results in UC-format
  
  uc_colnames <- c(
    "record_type", "cluster_number", 
    "length_size", "percent_similarity", 
    "match_orientation", "X6", "X7", 
    "X8", "X9", "X10")
  
  colnames(centroids_uc.tbl) <- uc_colnames # Assign informative column names to the UC-format clustering output
  
  metadata.tbl$n_clusters[i] <- 
    length(unique(centroids_uc.tbl$cluster_number)) # Record the number of clusters generated for the current sample
  
  # --------------------------
  # Prepare centroid database
  # --------------------------
  
  db.tbl <- centroids_uc.tbl %>% 
    filter(record_type == "S") %>%         # Retain centroid (seed) sequences from each predicted cluster
    rename(Header = X9) %>% 
    select(Header) %>% 
    left_join(bio_seqs.tbl, by = "Header") # Attach nucleotide sequences to centroid headers
  
  # -----------------------------
  # Map low-abundance sequences
  # -----------------------------
  
  if (nrow(noise_seqs.tbl) > 0) {
    noise_mapped.tbl <- vs_usearch_global(
      noise_seqs.tbl, 
      uc = TRUE,
      database = db.tbl,
      id = id,
      threads = 4) # Map low-abundance sequences against cluster centroids using the same identity threshold
    
    noise_colnames <- c(
      "record_type", "cluster_number", 
      "length", "percent_similarity", 
      "match_orientation", "X6", "X7", 
      "X8", "query", "target")
    
    colnames(noise_mapped.tbl) <- noise_colnames # Assign informative column names to the UC-format mapping output
  }
  
  # ---------------------------------
  # Parse initial clustering results
  # ---------------------------------
  
  c.tbl <- centroids_uc.tbl %>% 
    rename(
      Header = X9,
      centroid = X10) %>%                                        # Rename positional UC columns to meaningful names
    select(-c(
      "length_size", "match_orientation", "X6", "X7", "X8")) %>% # Remove UC columns not required for downstream analysis
    filter(record_type != "C") %>%                               # Remove cluster summary records
    mutate(size = str_extract(Header, "(?<=;size=)\\d+")) %>%    # Extract sequence abundance from the FASTA header
    mutate(size = as.numeric(size)) %>% 
    mutate(cluster_number = as.character(cluster_number)) %>% 
    mutate(centroid = ifelse(centroid == "*", Header, centroid)) # Replace the centroid placeholder with the sequence's own header for seed records
  
  # ----------------------------------
  # Parse mapped low-abundance sequences
  # ----------------------------------
  
  if (nrow(noise_seqs.tbl) > 0) {
    
    n.tbl <- noise_mapped.tbl %>% 
      rename(
        Header = query,
        centroid = target) %>%                                  # Rename query and target fields for consistency with the clustering results
      select(-c(
        "length", "match_orientation", "X6", "X7", "X8")) %>%   # Remove UC columns not required for downstream analysis
      mutate(size = str_extract(Header, "(?<=;size=)\\d+")) %>% # Parse sequence abundance from each query header
      mutate(size = as.numeric(size)) %>% 
      mutate(
        cluster_number = ifelse(
          cluster_number != "*", cluster_number, NA)) %>%       # Assign NA to sequences that could not be mapped to a cluster
      mutate(centroid = ifelse(centroid == "*", NA, centroid))  # Assign NA where no centroid match was identified
    
    metadata.tbl$n_reads_not_clustered[i] <- 
      sum(
        n.tbl %>% 
          filter(record_type == "N") %>% 
          select(size)) # Record the total number of reads represented by unmapped low-abundance sequences
    
    metadata.tbl$n_unique_reads_not_clustered[i] <-
      nrow(
        n.tbl %>% 
          filter(record_type == "N")) # Record the number of unique low-abundance sequences that remained unmapped
    
  } else {
    
    n.tbl <- tibble(
      record_type = character(),
      cluster_number = character(),
      percent_similarity = character(),
      Header = character(),
      centroid = character(),
      size = numeric()) # Create an empty table with the expected structure when no low-abundance sequences are present
    
    metadata.tbl$n_reads_not_clustered[i]        <- 0 # No reads remained unclustered
    metadata.tbl$n_unique_reads_not_clustered[i] <- 0 # No unique sequences remained unclustered
  }
  
  # Save n.tbl for later
  save(n.tbl, file = file.path(
    noise_seqs_cluster_size,
    paste0("noise_sequences_", metadata.tbl$sample_id[i], ".RData"))) # Save mapped and unmapped low-abundance sequences for downstream inspection
  
  # --------------------------------
  # Non-clustered species statistics
  # --------------------------------
  
  # Get metadata statistics for non-clustered species
  df1 <- n.tbl %>% 
    mutate(
      taxonomy = Header,
      taxonomy = str_match(taxonomy, "tax=(.*)_[0-9]+_[0-9]+/[0-9]")[,2]) # Extract ground truth species labels from low-abundance sequences
  
  df2 <- c.tbl %>% 
    mutate(
      taxonomy = Header,
      taxonomy = str_match(taxonomy, "tax=(.*)_[0-9]+_[0-9]+/[0-9]")[,2]) # Extract ground truth species labels from clustered sequences
  
  missing_in_hits <- anti_join(
    df1 %>% 
      filter(record_type == "N"), 
    bind_rows(df2, df1 %>% filter(record_type == "H")), 
    by = "taxonomy") # Identify species represented only by unmapped sequences and absent from all clustered or mapped sequences
  
  metadata.tbl$n_species_not_clustered[i] <- 
    n_distinct(missing_in_hits$taxonomy) # Record the number of species completely absent from the final clustering results
  
  # ----------------------------------
  # Combine all clustering assignments
  # ----------------------------------
  
  res.tbl <- bind_rows(c.tbl, n.tbl) %>%
    arrange(cluster_number) %>% 
    mutate(
      predicted_cluster = centroid,
      predicted_cluster = str_match(
        predicted_cluster, "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>% # Extract the taxonomic identity of the centroid as the predicted cluster identity
    mutate(
      actual_cluster = Header,
      actual_cluster = str_match(
        actual_cluster, "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>%    # Extract the ground truth species identity of each sequence
    left_join(all.tbl %>% select(Header, Sequence), by = "Header")    # Attach nucleotide sequences to clustering assignments
  
  # ---------------
  # Purity analysis
  # ---------------
  
  clustering_results.tbl <- res.tbl # Alias for readability in downstream clustering analyses
  
  cluster_summary <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number) %>%
    summarise(
      total_reads = sum(size),
      distinct_actual_clusters = 
        n_distinct(actual_cluster),    # Count number of distinct actual clusters in the predicted cluster
      .groups = "drop"
    ) %>%
    mutate(status = ifelse(            
      distinct_actual_clusters == 1,
      "Pure",
      "Mixed")) %>%                    # Classify clusters as pure (single species) or mixed (multiple species)
    arrange(status, desc(total_reads)) # Sort clusters by purity status and descending read count
  
  metadata.tbl$n_pure_clusters[i] <- 
    sum(cluster_summary$status == "Pure") # Record the number of pure clusters for the current sample
  
  purity_summary <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number, actual_cluster) %>%
    summarise(size_per_group = sum(size), 
              .groups = "drop_last") %>%              # Sum read counts for each species within each predicted cluster
    arrange(desc(size_per_group)) %>%
    summarise(
      total_cluster_size = sum(size_per_group),       # Total number of reads assigned to the predicted cluster
      dominant_actual = first(actual_cluster),        # Ground truth species contributing the largest number of reads
      dominant_size = first(size_per_group),          # Number of reads contributed by the dominant species
      purity_pct = dominant_size / total_cluster_size # Cluster purity: proportion of reads belonging to the dominant species
    ) %>%
    arrange(purity_pct)
  
  total_reads <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>% 
    select(size) %>% 
    sum() # Calculate the total number of reads assigned to predicted clusters
  
  metadata.tbl$average_purity_score[i] <- 
    mean(purity_summary$purity_pct) # Calculate mean purity across individual predicted clusters
  
  # -------------------------
  # Perfect clusters analysis
  # -------------------------
  
  input_totals <- all.tbl %>%
    group_by(actual_cluster) %>%
    summarise(
      total_size_in = sum(size, na.rm = TRUE), 
      .groups = "drop") # Calculate the total input read count for each species before clustering
  
  cluster_summary <- clustering_results.tbl %>%
    group_by(cluster_number) %>%
    summarise(
      n_species_in_cluster = n_distinct(actual_cluster), # Count the number of distinct species within each predicted cluster
      cluster_species = first(actual_cluster),           # Record species identity for single-species clusters
      total_size_out = sum(size, na.rm = TRUE),          # Calculate the total number of reads represented by each cluster
      .groups = "drop"
    )
  
  species_distribution <- clustering_results.tbl %>%
    group_by(actual_cluster) %>%
    summarise(
      n_groups_for_species = n_distinct(
        cluster_number, 
        na.rm = FALSE), # Count how many predicted clusters each ground truth species is distributed across
      .groups = "drop")
  
  perfect_clusters <- cluster_summary %>%
    filter(!is.na(cluster_number)) %>%
    filter(n_species_in_cluster == 1) %>%   # Retain only clusters containing a single species
    left_join(
      species_distribution, 
      by = c("cluster_species" = "actual_cluster")
    ) %>%
    filter(n_groups_for_species == 1) %>%   # Retain only species represented by exactly one predicted cluster
    left_join(
      input_totals, 
      by = c("cluster_species" = "actual_cluster")
    ) %>%
    filter(total_size_out == total_size_in) # Retain clusters containing all reads belonging to the corresponding species
  
  metadata.tbl$n_perfect_clusters[i] <- 
    nrow(perfect_clusters) # Record the number of perfect clusters: pure, non-split, and complete
  
  
  # ------------------------------------------
  # Represented species and abundance profiles
  # ------------------------------------------
  
  dominant_species_per_cluster <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number, actual_cluster) %>%
    summarise(
      total_size = sum(size), 
      .groups = "drop_last"
    ) %>%
    arrange(cluster_number, desc(total_size)) %>%
    summarise(
      dominant_actual_cluster = first(actual_cluster), 
      .groups = "drop") # Assign each predicted cluster to the species contributing the largest number of reads
  
  save(
    dominant_species_per_cluster,
    file = file.path(
      dominant_species_folder_cluster_size,
      paste0(
        "dominant_species_per_cluster_",
        metadata.tbl$sample_id[i],
        ".RData"))) # Save dominant-species assignments for downstream abundance analysis
  
  metadata.tbl$n_species_represented[i] <- 
    length(unique(dominant_species_per_cluster$dominant_actual_cluster)) # Record the number of species represented by at least one predicted cluster
  
  n_species_with_multiple_clusters <- dominant_species_per_cluster %>%
    group_by(dominant_actual_cluster) %>%
    summarise(
      n_clusters = n(), 
      .groups = "drop"
    ) %>%
    filter(n_clusters >= 2) %>%
    nrow() # Count species that dominate multiple predicted clusters (over-splitting)
  
  metadata.tbl$n_species_with_multiple_clusters[i] <- 
    n_species_with_multiple_clusters # Record the number of over-split species
  
  # ------------------
  # Abundance analysis
  # ------------------
  
  fq <- readFastq(
    file.path(fastq.folder, metadata.tbl$R1_file[i])
  ) %>%
    mutate(
      actual_cluster = Header,
      actual_cluster = str_match(
        actual_cluster, 
        "tax=(.*)_[0-9]+_[0-9]+/[0-9]")[,2]) # Read raw R1 reads and extract the ground truth species identity
  
  abundance_per_initiate_species <- fq %>%
    group_by(actual_cluster) %>%
    summarise(
      gold_standard_n = n(), 
      .groups = "drop"
    ) %>%
    mutate(
      gold_standard_rel_abundance = 
        gold_standard_n / sum(gold_standard_n)
    ) %>%
    arrange(desc(gold_standard_rel_abundance)) # Calculate ground truth relative abundance directly from raw read counts
  
  abundance_per_species <- clustering_results.tbl %>%
    left_join(
      dominant_species_per_cluster, 
      by = "cluster_number"
    ) %>% 
    mutate(
      dominant_actual_cluster = replace_na(
        dominant_actual_cluster, 
        "Unassigned"
      )
    ) %>%
    group_by(dominant_actual_cluster) %>%
    summarise(
      n = sum(size), 
      .groups = "drop"
    ) %>%
    mutate(rel_abundance = n / sum(n)) %>%
    arrange(desc(rel_abundance)) %>%
    rename(actual_cluster = dominant_actual_cluster) # Calculate observed species relative abundance from the clustering results
  
  abundance.tbl <- abundance_per_initiate_species %>%
    left_join(
      abundance_per_species, 
      by = "actual_cluster"
    ) %>%
    mutate(
      n = replace_na(n, 0),
      rel_abundance = replace_na(rel_abundance, 0)) # Combine ground truth and observed abundance profiles and assign zero to missing species
  
  save(
    abundance.tbl,
    file = file.path(
      abundance_folder_cluster_size,
      paste0(
        "abundance_",
        metadata.tbl$sample_id[i],
        ".RData"))) # Save abundance profiles for downstream comparisons
  
  metadata.tbl$abundance_correlation[i] <- cor(
    abundance.tbl$gold_standard_rel_abundance, 
    abundance.tbl$rel_abundance) # Calculate Pearson correlation between ground truth and observed relative abundances
}

save(
  metadata.tbl, 
  file = file.path(cluster_size_folder, "metadata.tbl.RData")) # Save metadata with all cluster_size performance statistics
```

## Plotting Preparations After Processing All Datasets

```{r}
# IMPORTANT:
# Run this section only after all datasets have been fully processed with the
# cluster_size pipeline. The code below loads the final metadata table from
# each dataset and combines them into a single table for downstream plotting
# and comparative analyses.

# --------------------------------
# Load per-dataset metadata tables
# --------------------------------

load("100_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_100s_10d.tbl <- metadata.tbl

load("100_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_100s_100d.tbl <- metadata.tbl

load("1000_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_1000s_10d.tbl <- metadata.tbl

load("1000_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_1000s_100d.tbl <- metadata.tbl                                        

load("10000_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_10000s_10d.tbl <- metadata.tbl                                     

load("10000_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_10000s_100d.tbl <- metadata.tbl                                        

# --------------------------------------------
# Combine datasets by simulated species richness
# --------------------------------------------

metadata_100s <- bind_rows(
  metadata_100s_10d.tbl,
  metadata_100s_100d.tbl) # Combine all 100-species datasets across dynamic ranges

metadata_1000s <- bind_rows(
  metadata_1000s_10d.tbl,
  metadata_1000s_100d.tbl) # Combine all 1000-species datasets across dynamic ranges

metadata_10000s <- bind_rows(
  metadata_10000s_10d.tbl,
  metadata_10000s_100d.tbl) # Combine all 10000-species datasets across dynamic ranges

# --------------------------------
# Combine all processed datasets
# --------------------------------

metadata_cluster_size_all.tbl <- bind_rows(
  metadata_100s,
  metadata_1000s,
  metadata_10000s) %>%
  mutate(pipeline = "cluster_size") # Add a pipeline identifier for downstream comparisons across clustering methods

save(
  metadata_cluster_size_all.tbl,
  file = "metadata_all_cluster_size_tbl.RData") # Save the combined metadata table for downstream plotting and comparative analyses
```

# UNOISE clustering

```{r}
minsize <- 8 # Minimum sequence abundance required for UNOISE clustering

load(file.path(rsearch_dir, "metadata.tbl.RData")) # Load metadata generated during read preprocessing

for (i in 1:nrow(metadata.tbl)) { # Iterate over each sample in the metadata table
  
  message(paste0("Processing sample ", 
                 metadata.tbl$sample_id[i], 
                 " (", i, "/", nrow(metadata.tbl), ")"))
  
  # ---------------------------
  # Load and parse input data
  # ---------------------------
  
  all.tbl <- readFasta(file.path(
    fasta.folder, metadata.tbl$dereplicated_fasta[i])) %>%                       # Read dereplicated sequences for current sample
    mutate(actual_cluster = Header,
           actual_cluster = str_match(actual_cluster, 
                                      "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>% # Extract the ground truth species label from each FASTA header
    mutate(size = str_extract(Header, "(?<=;size=)\\d+") %>% 
             as.numeric())                                                       # Parse sequence abundance (read count) from each dereplicated FASTA header
  
  metadata.tbl$n_species_pre_clustering[i] <- 
    length(unique(all.tbl$actual_cluster)) # Record the number of species represented immediately before clustering
  
  # -----------------
  # UNOISE clustering
  # -----------------
  
  centroids_unoise.tbl <- vs_cluster_unoise(
    all.tbl,
    uc = TRUE,
    minsize = minsize,
    threads = 4) # Perform UNOISE clustering and return sequence assignments in UC format
  
  uc_colnames <- c(
    "record_type", "cluster_number", 
    "length_size", "percent_similarity", 
    "match_orientation", "X6", "X7", 
    "X8", "X9", "X10")
  
  colnames(centroids_unoise.tbl) <- uc_colnames # Assign informative column names to the UC-format clustering output
  
  metadata.tbl$n_clusters[i] <- 
    length(unique(centroids_unoise.tbl$cluster_number)) # Record the number of clusters generated for the current sample
  
  # ------------------------
  # Parse clustering results
  # ------------------------
  
  c.tbl <- centroids_unoise.tbl %>% 
    rename(
      Header = X9,
      centroid = X10) %>%                                        # Rename positional UC columns to meaningful names
    select(-c(
      "length_size", "match_orientation", "X6", "X7", "X8")) %>% # Remove UC columns not required for downstream analysis
    filter(record_type != "C") %>%                               # Remove cluster summary records
    mutate(size = str_extract(Header, "(?<=;size=)\\d+")) %>%    # Parse sequence abundance from the FASTA header
    mutate(size = as.numeric(size)) %>% 
    mutate(cluster_number = as.character(cluster_number)) %>% 
    mutate(
      centroid = ifelse(centroid == "*", Header, centroid))      # Replace the centroid placeholder with the sequence's own header for seed records
  
  # --------------------------------
  # Identify non-clustered sequences
  # --------------------------------
  
  n.tbl <- all.tbl %>% 
    select(-c(actual_cluster, Sequence)) %>% 
    filter(!(Header %in% c.tbl$Header)) # Identify input sequences not represented in the UNOISE clustering output
  
  metadata.tbl$n_reads_not_clustered[i] <- 
    sum(n.tbl %>% select(size)) # Record the total number of reads represented by non-clustered sequences
  
  metadata.tbl$n_unique_reads_not_clustered[i] <- 
    nrow(n.tbl) # Record the number of unique sequences not represented in the clustering output
  
  n.tbl <- n.tbl %>% 
    mutate(record_type = "N",
           cluster_number = NA_character_,
           percent_similarity = "*",
           centroid = NA_character_) # Format non-clustered sequences for compatibility with the clustering results table
  
  save(n.tbl, file = file.path(
    noise_seqs_cluster_unoise,
    paste0("noise_sequences_", metadata.tbl$sample_id[i], ".RData"))) # Save non-clustered sequences for downstream inspection
  
  # ----------------------------------
  # Combine all clustering assignments
  # ----------------------------------
  
  res.tbl <- bind_rows(c.tbl, n.tbl) %>%
    arrange(cluster_number) %>% 
    mutate(
      predicted_cluster = centroid,
      predicted_cluster = str_match(
        predicted_cluster, "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>% # Extract the taxonomic identity of the centroid as the predicted cluster identity
    mutate(
      actual_cluster = Header,
      actual_cluster = str_match(
        actual_cluster, "tax=(.*)_[0-9]+_[0-9]+/[0-9]+;")[,2]) %>%    # Extract the ground truth species identity of each sequence
    left_join(all.tbl %>% select(Header, Sequence), by = "Header")    # Attach nucleotide sequences to clustering assignments
  
  # ---------------
  # Purity analysis
  # ---------------
  
  clustering_results.tbl <- res.tbl # Alias for readability in downstream clustering analyses
  
  cluster_summary <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number) %>%
    summarise(
      total_reads = sum(size),
      distinct_actual_clusters = 
        n_distinct(actual_cluster),    # Count the number of distinct ground truth species within each predicted cluster
      .groups = "drop"
    ) %>%
    mutate(status = ifelse(
      distinct_actual_clusters == 1,
      "Pure",
      "Mixed")) %>%                    # Classify clusters as pure (single species) or mixed (multiple species)
    arrange(status, desc(total_reads)) # Sort clusters by purity status and descending read count
  
  metadata.tbl$n_pure_clusters[i] <- 
    sum(cluster_summary$status == "Pure") # Record the number of pure clusters for the current sample
  
  purity_summary <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number, actual_cluster) %>%
    summarise(size_per_group = sum(size), .groups = "drop_last") %>% # Sum read counts for each species within each predicted cluster
    arrange(desc(size_per_group)) %>%
    summarise(
      total_cluster_size = sum(size_per_group),       # Total number of reads assigned to the predicted cluster
      dominant_actual = first(actual_cluster),        # Ground truth species contributing the largest number of reads
      dominant_size = first(size_per_group),          # Number of reads contributed by the dominant species
      purity_pct = dominant_size / total_cluster_size # Cluster purity: proportion of reads belonging to the dominant species
    ) %>%
    arrange(purity_pct)
  
  total_reads <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>% 
    select(size) %>% 
    sum() # Calculate the total number of reads assigned to predicted clusters
  
  metadata.tbl$average_purity_score[i] <- 
    mean(purity_summary$purity_pct) # Calculate mean purity across individual predicted clusters
  
  # -------------------------
  # Perfect clusters analysis
  #--------------------------
  
  input_totals <- all.tbl %>%
    group_by(actual_cluster) %>%
    summarise(total_size_in = sum(size, na.rm = TRUE), .groups = "drop") # Calculate the total input read count for each species before clustering
  
  cluster_summary <- clustering_results.tbl %>%
    group_by(cluster_number) %>%
    summarise(
      n_species_in_cluster = n_distinct(actual_cluster), # Count the number of distinct species within each predicted cluster
      cluster_species = first(actual_cluster),           # Record species identity for single-species clusters
      total_size_out = sum(size, na.rm = TRUE),          # Calculate the total number of reads represented by each cluster
      .groups = "drop"
    )
  
  species_distribution <- clustering_results.tbl %>%
    group_by(actual_cluster) %>%
    summarise(n_groups_for_species = n_distinct(cluster_number, na.rm = FALSE), 
              .groups = "drop") # Count how many predicted clusters each ground truth species is distributed across
  
  perfect_clusters <- cluster_summary %>%
    filter(!is.na(cluster_number)) %>%
    filter(n_species_in_cluster == 1) %>%   # Retain only clusters containing a single species
    left_join(
      species_distribution, 
      by = c("cluster_species" = "actual_cluster")
    ) %>%
    filter(n_groups_for_species == 1) %>%   # Retain only species represented by exactly one predicted cluster
    left_join(
      input_totals, 
      by = c("cluster_species" = "actual_cluster")
    ) %>%
    filter(total_size_out == total_size_in) # Retain clusters containing all reads belonging to the corresponding species
  
  metadata.tbl$n_perfect_clusters[i] <- 
    nrow(perfect_clusters) # Record the number of perfect clusters: pure, non-split, and complete
  
  
  # ------------------------------------------
  # Represented species and abundance profiles
  #-------------------------------------------
  
  dominant_species_per_cluster <- clustering_results.tbl %>%
    filter(!is.na(cluster_number)) %>%
    group_by(cluster_number, actual_cluster) %>%
    summarise(total_size = sum(size), .groups = "drop_last") %>%
    arrange(cluster_number, desc(total_size)) %>%
    summarise(dominant_actual_cluster = first(actual_cluster), .groups = "drop") # Assign each predicted cluster to the species contributing the largest number of reads
  
  save(dominant_species_per_cluster,
       file = file.path(
         dominant_species_folder_cluster_unoise,
         paste0("dominant_species_per_cluster_", 
                metadata.tbl$sample_id[i], ".RData"))) # Save dominant-species assignments for downstream abundance analysis
  
  metadata.tbl$n_species_represented[i] <- 
    length(unique(dominant_species_per_cluster$dominant_actual_cluster)) # Record the number of species represented by at least one predicted cluster
  
  n_species_with_multiple_clusters <- dominant_species_per_cluster %>%
    group_by(dominant_actual_cluster) %>%
    summarise(
      n_clusters = n(), 
      .groups = "drop"
    ) %>%
    filter(n_clusters >= 2) %>%
    nrow() # Count species that dominate multiple predicted clusters (over-splitting)
  
  metadata.tbl$n_species_with_multiple_clusters[i] <- 
    n_species_with_multiple_clusters # Record the number of over-split species
  
  # ------------------
  # Abundance analysis
  # ------------------
  
  fq <- readFastq(file.path(fastq.folder, metadata.tbl$R1_file[i])) %>%
    mutate(
      actual_cluster = Header,
      actual_cluster = str_match(
        actual_cluster, "tax=(.*)_[0-9]+_[0-9]+/[0-9]")[,2]) # Read raw R1 reads and extract the ground truth species identity
  
  abundance_per_initiate_species <- fq %>%
    group_by(actual_cluster) %>%
    summarise(gold_standard_n = n(), .groups = "drop") %>%
    mutate(gold_standard_rel_abundance = gold_standard_n / sum(gold_standard_n)) %>%
    arrange(desc(gold_standard_rel_abundance)) # Calculate ground truth relative abundance directly from raw read counts
  
  abundance_per_species <- clustering_results.tbl %>%
    left_join(dominant_species_per_cluster, by = "cluster_number") %>% 
    mutate(
      dominant_actual_cluster = replace_na(
        dominant_actual_cluster, "Unassigned")) %>%
    group_by(dominant_actual_cluster) %>%
    summarise(n = sum(size), .groups = "drop") %>%
    mutate(rel_abundance = n / sum(n)) %>%
    arrange(desc(rel_abundance)) %>%
    rename(actual_cluster = dominant_actual_cluster) # Calculate observed species relative abundance from the clustering results
  
  abundance.tbl <- abundance_per_initiate_species %>%
    left_join(abundance_per_species, by = "actual_cluster") %>%
    mutate(n = replace_na(n, 0),
           rel_abundance = replace_na(rel_abundance, 0)) # Combine ground truth and observed abundance profiles and assign zero to missing species
  
  save(abundance.tbl,
       file = file.path(
         dominant_species_folder_cluster_unoise,
         paste0("abundance_", metadata.tbl$sample_id[i], ".RData"))) # Save abundance profiles for downstream comparisons
  
  metadata.tbl$abundance_correlation[i] <- cor(
    abundance.tbl$gold_standard_rel_abundance, 
    abundance.tbl$rel_abundance) # Calculate Pearson correlation between ground truth and observed relative abundances
}

save(
  metadata.tbl, 
  file = file.path(cluster_unoise_folder, "metadata.tbl.RData")) # Save metadata with all UNOISE clustering performance statistics
```

## Plotting Preparations After Processing All Datasets

```{r}
# IMPORTANT:
# Run this section only after all datasets have been fully processed with the
# UNOISE pipeline using minsize = 8. The code below loads the final metadata
# table from each dataset and combines them into a single table for downstream
# plotting and comparative analyses.

# --------------------------------
# Load per-dataset metadata tables
# --------------------------------

load("100_species_10_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_100s_10d.tbl <- metadata.tbl            

load("100_species_100_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_100s_100d.tbl <- metadata.tbl                                          

load("1000_species_10_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_1000s_10d.tbl <- metadata.tbl                                        

load("1000_species_100_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_1000s_100d.tbl <- metadata.tbl                                       

load("10000_species_10_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_10000s_10d.tbl <- metadata.tbl                                       

load("10000_species_100_dynamic/rsearch_results/cluster_unoise_8/metadata.tbl.RData")
metadata_10000s_100d.tbl <- metadata.tbl                                     

# --------------------------------
# Load per-dataset metadata tables
# --------------------------------

load("100_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_100s_10d.tbl <- metadata.tbl

load("100_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_100s_100d.tbl <- metadata.tbl

load("1000_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_1000s_10d.tbl <- metadata.tbl

load("1000_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_1000s_100d.tbl <- metadata.tbl                                        

load("10000_species_10_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_10000s_10d.tbl <- metadata.tbl                                     

load("10000_species_100_dynamic/rsearch_results/cluster_size/metadata.tbl.RData")
metadata_10000s_100d.tbl <- metadata.tbl                                        

# --------------------------------------------
# Combine datasets by simulated species richness
# --------------------------------------------

metadata_100s <- bind_rows(
  metadata_100s_10d.tbl,
  metadata_100s_100d.tbl) # Combine all 100-species datasets across dynamic ranges

metadata_1000s <- bind_rows(
  metadata_1000s_10d.tbl,
  metadata_1000s_100d.tbl) # Combine all 1000-species datasets across dynamic ranges

metadata_10000s <- bind_rows(
  metadata_10000s_10d.tbl,
  metadata_10000s_100d.tbl) # Combine all 10000-species datasets across dynamic ranges

# --------------------------------
# Combine all processed datasets
# --------------------------------

metadata_cluster_unoise_8_all.tbl <- bind_rows(
  metadata_100s,
  metadata_1000s,
  metadata_10000s) %>%
  mutate(pipeline = "cluster_unoise_8") # Add a pipeline identifier for downstream comparisons across clustering methods

save(
  metadata_cluster_unoise_8_all.tbl,
  file = "metadata_all_cluster_unoise_8_tbl.RData") # Save the combined metadata table for downstream plotting and comparative analyses

